\begin{abstract} 

Phase Change Memory (PCM) is an upcoming \emph{nonvolatile} memory
technology that is comparable to traditional DRAM with regard to read
latency, and markedly superior with regard to storage density and 
idle power consumption. Due to these desirable characteristics, PCM is expected
to play a significant role in the next generation of computing systems.
However, it also has limitations in the form of expensive writes and
limited write endurance. Correspondingly, there has been recent research
investigating how database engines may be redesigned to suit
DBMS deployments on the new technology.

In this paper, we propose the design of PCM-conscious
database operators for the model of PCM deployment in which PCM augmented with a small DRAM buffer acts as the main memory.  Specifically,
we propose novel implementations of the ``workhorse" database operators:
\textit{sort}, \textit{hash join} and \textit{group-by} that reduce both writes as well as query response times. We also provide
estimators of the writes incurred by these techniques. 

Then, we
incorporate the proposed techniques in a state-of-the-art architectural
simulator and assess their performance on TPC-H benchmark queries. The estimators are validated against the actual writes obtained during experiments.
The experimental results suggest that their collective impact can increase
the PCM lifetime by upto four times while simultaneously improving the
query response times. In essence, our algorithms provide both short term and long term improvements. These outcomes augur well for database engines
that wish to leverage the impending transition to PCM-based computing.
\end{abstract}
